\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-02-05 document/XSGRB_sample.tex
Final draft sent around

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph up
	to \termdate. This totals spectroscopic observations of 103 individual bursts
	observed with 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneous sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of the
	underlying host galaxies. We provide an assessment of the degree of completeness
	with respect to the parent GRB population, in terms of the X-ray properties of
	the bursts in the sample and find that the sample presented here is
	representative of the full \textit{Swift} sample. For all bursts (42) for which
	it is possible, we provide a measurement of the neutral hydrogen column density,
	increasing the total number of published H{\sc I} column density measurements by
	$\sim$ 50 per cent. %	We
	% constrain the
	% fraction
	%	of dark bursts to be less than 28 per cent based on the detection of an
	% optical
	%	afterglow and tentatively confirm previous results that larger optical
	% darkness
	%	is correlated with increased X-ray absorption.
	This dataset provides a unique resource to study the ISM across cosmic time,
	from the local progenitor surroundings to the galaxy-scale conditions.}

@@

$\gamma$-ray bursts (GRBs) provide constraints on a very wide range of topics in
astrophysics. Examples range from small-scale phenomena relating to magnetars,
properties of highly relativistic jets, hyper/supernova explosions, the
interstellar medium, dust extinction curves, starburst galaxies, chemical and
molecular abundances, escape of ionizing radiation, the ionization state of the
intergalactic medium, intervening absoption systems to standard candles in
cosmology \citep[e.g.,][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

@@

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=8cm]{figures/resolutions}} \caption{Blue
	datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric
	absorption lines in the VIS spectra, as a function of the airmass corrected
	ambient FWHM. The red datapoints show a subsample of measurements obtained for
	NIR spectra. The colored lines show the corresponding linear fits to the
	datapoints. The colored bounds contain 68 per cent of the best-fit probability
	mass. } \label{fig:res}
\end{figure}

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline.

%First we fit a Gaussian function
%in the spatial direction of the trace of the standard star at 792 nm (i.e. in
%the VIS arm). After this, we fit a series of  20 telluric absorption lines in
%the telluric standard star spectra with Gaussians, taking care to select
%transitions that are not almost-resolved multiples, should be intrinsically
%unresolved, and are in areas with well defined continuum flux. We pick 34
%telluric standard stars spanning a range of DIMM seeing values, with the
%majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
%(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
%linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
%sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$ (see Fig.
%\ref{fig:res}). We use this linear relation as a way to estimate the spectral
%resolution for medium to low signal to noise afterglow spectra in the VIS arm.
%To extend this to the UVB and NIR arm, we measured a series of lines in the NIR
%arm spectra of a subset of 19 sources used for the VIS arm above, and find that
%the resulting distribution is consistent with a simple scaling of the VIS arm
%relation by the ratio of resolutions of the NIR and VIS arm for unresolved, slit
%filling, sources as given on the ESO instrument website. The UVB arm contains no
%suitable absorption lines to use, and we therefore use a scaled value as in the
%NIR arm. This simple analysis gives a sufficiently accurate estimate for the
%analysis of the low signal-to-noise science spectra.

%fit parameter 1-sigma error
%———————————-
%6.96733427249 +- 1.35492333728
%36.1113577532 +- 1.36818568362
%fit parameter 1-sigma error
%———————————-
%3.64773146238 +- 0.563740522948
%24.2308362925 +- 0.792580030602

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption lines
imprinted on the GRB afterglow continuum. Because the optical afterglow rapidly
fades (typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations
for any observation is for the short, $z = 1.717$, GRB~160410A for which
spectroscopic integration was initiated only 8.4 minutes after the BAT trigger.
To illustrate the importance of the follow-up delay for the redshift
completeness, we plot the redshift completeness as a function of delay time in
Fig. \ref{fig:timing} for all the bursts we have followed up, including the ones
outside the statistical sample. As can be seen from the figure, the fraction of
GRBs with a redshift determination decreases with follow-up delay. The redshift
completeness for bursts that we have followed up is 92 per cent. This degree of
completeness in the followed bursts, illustrates the efficiency of VLT/X-shooter
in redshift determination. Not shown in the figure are an additional 12 bursts
that have redshift determinations based on late-time host observations with
delay times longer than $\sim$10 days.

@@

Of all the BAT-triggered bursts, a total of 158 bursts fulfill the sample
criteria specified in Sect. \ref{samplecrit}, since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 61 per cent. We return to the
question of redshift completeness in Sect. \ref{redshift}. From this sample, 91
GRBs have been spectroscopically followed up with X-shooter. In order to assess
whether the subset of bursts followed up are representative of the underlying
GRB parent population, we compare intrinsic properties of GRBs in our sample to
GRBs in the full sample followed up by \textit{Swift}. We show the comparison
between the BAT (15 -- 150 keV) fluence, the XRT flux (0.3 -- 10 keV) at 11
hours, and the intrinsic X-ray derived equivalent hydrogen column density at the
redshift of the GRB, in excess of the Galactic X-ray absorption column,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the full \swift~sample. We return to
the last point in Sect. \ref{darkness}.

@@

Out of the 158 bursts in the statistical sample, 33 ( $\sim$ 20 per cent) were
not observed due to reasons unrelated to the GRB or afterglow properties. The
reasons include unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visiting observer rejecting the ToO trigger, or bad
weather. Because this cut is unrelated to the GRB properties, it will not change
the statistical properties of the full sample. Removing these bursts for the
statistical sample, dramatically improves the redshift completeness from 61 per cent to 84
per cent. The remaining burst not followed up already had redshift from other
instruments or were very faint and without a host association thus observations
unlikely to yield a redshift measurement. In the remainder of the text, we
consider the 125 bursts our statistical sample.

@@

Because the redshift completeness of our statistical sample is 84 per cent,
making inference of the \textit{true} redshift distribution of GRBs based on
this sample is impossible. %The biases introduced in the measured redshift
%distribution by our increased ability to secure redshifts for optically
% brighter
%bursts has been investigated by \citet{Turpin2016} who find that we
%systematically are able to measure redshifts for longer duration GRBs.
%Additionally o
For instance, only the brightest GRBs are seen above redshift $z \gtrsim 1$ as
shown in Fig. \ref{fig:z}. As described in detail in \citet{Hjorth2012} and
\citet{Perley2016b}, bursts for which the redshift is measured from the
afterglow are systematically found for host galaxies with a lower luminosity
than burst for which the redshift is measured from the host galaxy. Only a few
GRBs hosted in galaxies, with stellar masses more than
$10^{10}~\mathrm{M}_\odot$ have the redshift measured based on the afterglow
continuum. This is likely related to the presence of higher contents of dust in
more massive galaxies, leading to a larger fraction of extincted afterglows.

@@

However, the fraction of dark bursts for which we have measured redshifts for is
lower than the for the ones with a detected optical afterglow. For the
afterglows we have observed as part of the statistical sample that do not have a
detected optical afterglow, the redshift completeness is 53 per cent. For
comparison, for the afterglows in the statistical sample we have observed with
an optical afterglow detected, the redshift completeness is 91 per cent. This
also goes to show that the lack of redshift completeness in the sample presented
here is in part due to the increased difficulty of measuring a redshift for
bursts without an afterglow. To measure a redshift, we either need a detected
afterglow to obtain a spectrum or to locate the host galaxy and determine the
redshift from there. It is more difficult to correctly associate a galaxy with a
burst when there is no detected optical afterglow and hence a correct redshift
measurement is more difficult to make, see \citet{Jakobsson2005, Levesque2010}
and \citet{Perley2017}.

@@

In the compilation of $N_{\mathrm{HI}}$ measurements towards GRBs in Tanvir et
al. (submitted) %\citet{Tanvir2017}
there are 81 published $N_{\mathrm{HI}}$ values, excluding the measurements
provided here. We here provide 39 new neutral hydrogen column density
measurements -- an increase of the number of optically derived hydrogen column
densities of $\sim$ 50 per cent. We show the two distributions in  Fig.
\ref{fig:NH_dist}. We compare the median, the 14th, and 86th percentiles of the
two distributions. The sample presented has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the
rest of the literature values has \nh~$= 21.5_{-1.5}^{+0.4}$. We see that the
two distributions have a large degree of overlap due to the large width of the
distributions, but we find a slightly higher median value for the new sample
presented here. A 2-sided KS test gives a p-value of p = 0.006, meaning
relatively large evidence against the null that the two samples are drawn from
the same underlying distribution. Because the bursts that have measurements of
the hydrogen column density are selected solely based on our ability to infer a
column, it is difficult to make any strong conclusions about the population
statistics in terms of gas content.

@@

The sample presented here includes spectroscopic observations of 91 systems
fulfilling our sample criteria, including 18 spectra that are late-time
observations of the underlying host galaxies. All spectra have been made
publicly available in the reduced form used in this paper.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectrographs. GRB afterglow sight-lines are unique in the
sense that only after observing more than 12000 damped Lyman-$\alpha$ absorbers
(DLAs) towards about 10$^5$ quasars, a handful systems with \nh~$ > 22$ been
identified \citep[e.g., five in][]{Noterdaeme2012b}. Long GRB afterglow spectra,
by contrast, reveal such systems in the majority of cases \citep[][and this
work]{Jakobsson2006b, Fynbo2009, Cucchiara2015}. With afterglow spectroscopy
(throughout the electromagnetic spectrum from X-rays to the sub-mm) we are able
to characterize the properties of star-forming galaxies over cosmic history in
terms of redshifts, metallicities, molecular contents, ISM temperatures, UV-flux
densities, extinction curves, etc.  A number of independent papers have been
published or submitted for publication focusing on many of these specific issues
of our sample such as extinction curves \citep[][see also \citealt{Fynbo2014,
	Heintz2017}]{Japelj2015, Zafar17}, emission lines from the underlying host
galaxies \citep{Kruhler2015}, the frequency of intervening \mgii~ absorbers
\citep{Christensen2017}, \citet{Arabsalmani2018} on the metallicity-scaling
relations, and escape of ionizing radiation (Tanvir et al., submitted). A number
of additional companion papers are also planned, investigating the detailed
properties of the sample presented here, including equivalent width
distributions (de Ugarte Postigo et al., in preparation), metallicities and
kinematics (Th{\"o}ne et al., in preparation), high ionization lines (Heintz et
al., in preparation), molecular lines (Bolmer et al., in preparation),
fine-structure lines (Vreeswijk et al., in preparation), composite GRB afterglow
spectrum (Selsing et al., in preparation).

The potential of using GRB sightlines as probes is far from fully harvested. The
sample of sightlines probed by our spectra are not representative for all GRB
sightlines as we have shown and consistent with earlier findings from samples
based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009} and from studies
of complete samples of GRB host galaxies \citep{Hjorth2012, Covino2013,
	Perley2016}. \cite{Kruhler2013} argue, that very rich sightlines like that
probed by the remarkable GRB~080607 \citep{Prochaska2009} are probably
significantly more frequent than in the sightlines sampled by our spectra.
However, with current instrumentation, these sightlines are out of reach except
under very fortunate circumstances as in the case of GRB~080607 when the
afterglow could be observed only a few minutes after the burst with a 10-m class
telescope. Observations of such sightlines with X-shooter-like spectrographs on
the next generation of 20--40-m telescopes is likely to be very rewarding, given
that a suitable GRB detector will be available.

@@

These observations have not been published before. Starting 12.36 hours after
the BAT trigger, 4.8 ks spectroscopic integration yields a very faint trace down
to $\sim$4300\AA, only visible after binning heavily. This could indicate a
redshift, $z ~\sim 2.5$, but no emission lines or absorption lines are
immediately visible to support this.

\subsection{GRB~110709B (z=2.109 (NEW!!!!))} \label{110709}

@@

This is a Fermi burst with a LAT detection and thus outside the statistical
sample, but nonetheless followed up due to the extremely high peak energy
\citep{Axelsson2012}. Starting 28.7 hours after the burst trigger, 2.4ks
spectroscopic observation reveals after heavy binning, a wide, faint trace down
to $\sim$ 5800 \AA, offset by 2.5\arcsec~relative to the centering of the slit.
No good redshift measurement can be inferred from this. We have adopted the
redshift from GCN  12193 \citep{GCN12193}

@@

Starting 6.15 hours after the BAT trigger, spectroscopic integration for 4.8ks
reveals a moderate signal-to-noise GRB afterglow continuum, down to $\sim$ 5000
\AA. The simultaneous detection of absorption features identified as \lya,
\SIii, \civ, \alii, \cah, \cak, and \mgii, and emission from the \oiii-doublet,
securely sets $z = 3.36$ as the redshift of the GRB. This data forms the basis
of GCN 12284 \citep{GCN12284}, but is not published elsewhere.

@@

This data formed the basis of GCN 12431 \citep{GCN12431} and is additionally
published in \citet{Sparre2014}. Observations of this GRB afterglow was
initiated 7.71 hours after the BAT trigger and had a duration of 8.4ks. A second
observational epoch started 20.1 hours and lasted for 6.6ks. The GRB afterglow
continuum is well detected down to $\sim$7600 \AA, with several strong
absorption features imprinted, all at a common $z = 4.990$. \lya~is clearly
detected and we additionally detect lines identified as \SIii, \feii, \civ,
\mgii, \SIii*, \sii*, \oi*. An intervening DLA systems is additionally detected
at  $z = 4.61$ as seen from \lya~and \mgii~absorption.

@@

The data presented here has been examined by \citet{Japelj2015} and
\citet{Zafar2017} who both find a significant amount of extinction, $A_V\approx
1$ mag. Three epochs of observations have been obtained, the first two
immediately after the burst, and the last one long after the afterglow had
faded. Starting 1.4 hours after the \textit{Swift} trigger, the first epoch
contains bright afterglow continuum. Rich in absorption features belonging to a
multitude of ions at $z =    1.728$ is estimated for the host with intervening
systems at $z = 1.476$, $z = 1.214$, $z = 0.662$ and $z = 0.632$. The second
epoch, obtained 4.5 hours after the burst contains the fading afterglow. A third
epoch is obtained $>1$ year after the GRB in which emission lines from \hb~and
\ha~are found at the redshift of the host, confirming the association of the
absorption line system and the host. We also detect C\,\textsc{i} in absorption
which indicates the presence of cold gas.

@@

\subsection{GRB~120224A (z = 1.10 NEW!!!)} \label{120224}

@@

\subsection{GRB~120311 (z = 0.350 NEW!!!)} \label{120311}

@@

\subsection{GRB~120805A (z = 3.9 NEW!!!)} \label{120805}

@@

This spectrum has not previously been published. Starting only 45 minutes after
the BAT trigger, $2\times600$ s spectroscopic integration carried on into the
beginning twilight. Observed at very high airmass with a quickly varying
background, a faint afterglow trace is visible across all arms of X-shooter,
down to ~ 4800 \AA, which if interpreted as the break due to \lya suggests $z =
2.9$. This supports the approximate redshift suggested in GCN 14898
\citep{GCN14898}.

@@

This data formed the basis for 15330 \citep{GCN15330}, but is not published
elsewhere. Starting $\sim$1.5 days after the Fermi-GBM trigger, 4.5 ks
spectroscopic integration captures a modest signal-to-noise GRB afterglow
continuum all the way down to $\sim$3200 \AA. Imprinted on the continuum are
absorption features, which we identify as due to \lya, \feii,  \mgii, \mgi~at
the same redshift, which we measure to be $z = 1.874$. From the broad absorption
trough due to \lya, we infer \nh~$= 22.0 \pm 0.3$. This spectrum is taken under
ESO programme ID: 092.D-0056(A) (PI: Rau).

\subsection{(CHECK ME - Bright un-used spectrum)GRB~131030A (z = 1.296)}	
\label{131030}

@@

\subsection{CHECK ME - Bright un-used spectrum - GRB~140311A (z = 4.954)}	
\label{140311}

>>>>

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. This totals spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneous sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results, that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

@@

$\gamma$-ray bursts (GRBs) are bright glimpses of electromagnetic radiation that
pierce through the universe, all the way from the beginning of time. They
provide constraints on a very wide range of topics in astrophysics. Examples
range from small-scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

@@

\begin{figure}[!ht]
	\centerline{\includegraphics[width=8cm]{figures/resolutuions}} \caption{Blue
		datapoints show the FWHM (km/s) of Gaussian fits to unresolved telluric
		absorption lines in the VIS spectra, as a function of the airmass corrected
		ambient FWHM. The red datapoints show a subsample of measurements obtained for
		NIR spectra. The colored lines show the corresponding linear fits to the
		datapoints. The colored bounds contain 68 per cent of the best-fit probability
		mass. The red and blue dashed line indicates the nominal resolution for the
		slit-widths used for the telluric standard observations. As can be see, the
		delivered resolution is in many cases superior to the nominal one. }
	\label{fig:res}
\end{figure}

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. To get the resolutions, we
select a series of atmospheric transitions that are resolved multiples which
should be intrinsically unresolved, and are in areas with well defined continuum
flux. We then fit the lines with Voigt-profiles and calculate a corresponding
FWHM, which we can then convert to a delivered resolution using the wavelength
of the chosen transitions. To get the resolution as a function of ambient
conditions, for each observations we also calculate the airmass-corrected
\href{http://www.eso.org/asm/ui/publicLog}{DIMM} seeing, which is measured at
500 nm.

The resulting distribution of spectral FWHM (km/s) as a function of spatial
FWHM at 500 nm is fairly well described in the VIS arm by a linear relation $a
+ b*x$, with $x$ the spatial FWHM in arcseconds, $a= 25.2 \pm 0.8$ km/s, $b=3.8
\pm 0.7$ (see Fig. \ref{fig:res}). For the NIR arm the corresponding relation
is $a= 36.0 \pm 1.8$ km/s, $b=8.4 \pm 1.8$. We use these linear relations as a
way to estimate the spectral resolution for medium to low signal-to-noise
afterglow spectra. To extend this to the UVB arm, we calculate the ratio
between the VIS and the NIR arm resolutions and find that the resulting
distribution is consistent with a simple scaling of the VIS arm relation by the
ratio of resolutions of the NIR and VIS arm for unresolved, slit filling,
sources as given on the
\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/inst.html}{ESO instrument website}. The UVB arm contains no suitable absorption lines to use, and we therefore use a scaled value. This simple analysis gives a sufficiently accurate estimate for the analysis of the low signal-to-noise science spectra.

@@

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption lines
imprinted on the GRB afterglow continuum. Because the optical afterglow rapidly
fades (typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations
for any observation is for the short, $z = 1.717$, GRB~160410A for which
spectroscopic integration was initiated only 8.4 minutes after the BAT trigger.
To illustrate the importance of the follow-up delay for the redshift
completeness, we plot the redshift completeness as a function of delay time in
Fig. \ref{fig:timing} for all the bursts we have followed up, including the ones
outside the statistical sample. As can be seen from the figure, the fraction of
GRBs with a redshift determination decreases with follow-up delay. The redshift
completeness for bursts that we have followed up is 94 per cent. This degree of
completeness in the followed bursts, illustrates the efficiency of VLT/X-shooter
in redshift determination. Not shown in the figure are an additional 12 bursts
that have redshift determinations based on late-time host observations with
delay times longer than $\sim$10 days.

@@

Of all the BAT-triggered bursts, a total of 165 bursts fulfill the sample
criteria specified in Sect. \ref{samplecrit}, since the commissioning of
VLT/X-shooter. This sample constitutes the \textit{statistical sample} from
which we will derive statistical properties of the GRB population. The redshift
completeness of the full statistical sample is 61 per cent. We return to the
question of redshift completeness in Sect. \ref{redshift}. From this sample, 93
GRBs have been spectroscopically followed up with X-shooter. In order to assess
whether the subset of bursts followed up are representative of the underlying
GRB parent population, we compare intrinsic properties of GRBs in our sample to
GRBs in the full sample followed up by \textit{Swift}. We show the comparison
between the BAT (15 -- 150 keV) fluence, the XRT flux (0.3 -- 10 keV) at 11
hours, and the intrinsic X-ray derived equivalent hydrogen column density at the
redshift of the GRB, in excess of the Galactic X-ray absorption column,
$N_{\mathrm{HI,X}}$, in Fig. \ref{fig:swift_complete}. For the latter, we can
only use values of $N_{\mathrm{HI,X}}$ derived for bursts with a measured
redshift, excluding $\sim 75$ per cent of the full \swift~sample. We return to
the last point in Sect. \ref{darkness}.

@@

Out of the 165 bursts in the statistical sample, 36 (22 per cent) were
not observed due to reasons unrelated to the GRB or afterglow properties. The
reasons include unavailability of the telescope due to technical maintenance
(e.g., mirror re-coating), a visiting observer rejecting the ToO trigger, or bad
weather. Because this cut is unrelated to the GRB properties, it will not change
the statistical properties of the full sample. Removing these bursts for the
statistical sample, dramatically improves the redshift completeness from 61 per cent to 88
per cent. The remaining burst not followed up already had redshift from other
instruments or were very faint and without a host association thus observations
unlikely to yield a redshift measurement. In the remainder of the text, we
consider the 129 bursts our statistical sample.

@@

Because the redshift completeness of our statistical sample is 88 per cent,
making inference of the \textit{true} redshift distribution of GRBs based on
this sample is impossible. %The biases introduced in the measured redshift
%distribution by our increased ability to secure redshifts for optically
% brighter
%bursts has been investigated by \citet{Turpin2016} who find that we
%systematically are able to measure redshifts for longer duration GRBs.
%Additionally o
For instance, only the brightest GRBs are seen above redshift $z \gtrsim 1$ as
shown in Fig. \ref{fig:z}. As described in detail in \citet{Hjorth2012} and
\citet{Perley2016b}, bursts for which the redshift is measured from the
afterglow are systematically found for host galaxies with a lower luminosity
than burst for which the redshift is measured from the host galaxy. Only a few
GRBs hosted in galaxies, with stellar masses more than
$10^{10}~\mathrm{M}_\odot$ have the redshift measured based on the afterglow
continuum. This is likely related to the presence of higher contents of dust in
more massive galaxies, leading to a larger fraction of extincted afterglows.

@@

However, the fraction of dark bursts for which we have measured redshifts for is
lower than the for the ones with a detected optical afterglow. For the
afterglows we have observed as part of the statistical sample that do not have a
detected optical afterglow, the redshift completeness is 53 per cent. For
comparison, for the afterglows in the statistical sample we have observed with
an optical afterglow detected, the redshift completeness is 92 per cent. This
also goes to show that the lack of redshift completeness in the sample presented
here is in part due to the increased difficulty of measuring a redshift for
bursts without an afterglow. To measure a redshift, we either need a detected
afterglow to obtain a spectrum or to locate the host galaxy and determine the
redshift from there. It is more difficult to correctly associate a galaxy with a
burst when there is no detected optical afterglow and hence a correct redshift
measurement is more difficult to make, see \citet{Jakobsson2005, Levesque2010}
and \citet{Perley2017}.

@@

12 of the $N_{\mathrm{HI}}$ measurements for these spectra have previously been
presented in \citet{Cucchiara2015} (See Table \ref{tab:HI}). We provide new
measurements here for completeness. In the compilation of $N_{\mathrm{HI}}$
measurements towards GRBs in Tanvir et al. (submitted) there are 93 published
$N_{\mathrm{HI}}$ values, excluding the measurements provided here. We here
provide 30 new neutral hydrogen column density measurements -- an increase of
the number of optically derived hydrogen column densities of $\sim$ 33 per cent.
We show the two distributions in  Fig. \ref{fig:NH_dist}. We compare the median,
the 14th, and 86th percentiles of the two distributions. The sample presented
has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the rest of the literature values has
\nh~$= 21.5_{-1.5}^{+0.4}$. We see that the two distributions have a large
degree of overlap due to the large width of the distributions, but we find a
slightly higher median value for the new sample presented here. A 2-sided KS
test gives a p-value of p = 0.006, meaning relatively large evidence against the
null that the two samples are drawn from the same underlying distribution.
Because the bursts that have measurements of the hydrogen column density are
selected solely based on our ability to infer a column, it is difficult to make
any strong conclusions about the population statistics in terms of gas content.

@@

The sample presented here includes spectroscopic observations of 93 systems
fulfilling our sample criteria, including 18 spectra that are late-time
observations of the underlying host galaxies. All spectra have been made
publicly available in the reduced form used in this paper.

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectrographs. GRB afterglow sight-lines are unique in the
sense that only after observing more than 12000 damped Lyman-$\alpha$ absorbers
(DLAs) towards about 10$^5$ quasars, a handful systems with \nh~$ > 22$ been
identified \citep[e.g., five in][]{Noterdaeme2012b}. Long GRB afterglow spectra,
by contrast, reveal such systems in the majority of cases \citep[][and this
work]{Jakobsson2006b, Fynbo2009, Cucchiara2015}. With afterglow spectroscopy
(throughout the electromagnetic spectrum from X-rays to the sub-mm) we are able
to characterize the properties of star-forming galaxies over cosmic history in
terms of redshifts, metallicities, molecular contents, ISM temperatures, UV-flux
densities, extinction curves, etc.  A number of independent papers have been
published or submitted for publication focusing on many of these specific issues
of our sample such as extinction curves (\citealt{Japelj2015}, Zafar et al.
submitted, see also \citealt{Fynbo2014, Heintz2017}]), emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\mgii~ absorbers \citep{Christensen2017}, \citet{Arabsalmani2018} on the
metallicity-scaling relations, and escape of ionizing radiation (Tanvir et al.,
submitted). A number of additional companion papers are also planned,
investigating the detailed properties of the sample presented here, including
equivalent width distributions (de Ugarte Postigo et al., in preparation),
metallicities and kinematics (Th{\"o}ne et al., in preparation), high ionization
lines (Heintz et al., in submitted), molecular lines (Bolmer et al., in
preparation), fine-structure lines (Vreeswijk et al., in preparation), and
composite GRB afterglow spectrum (Selsing et al., in preparation).

The potential of using GRB sightlines as probes is far from fully harvested. The
sample of sightlines probed by our spectra are not representative for all GRB
sightlines as we have shown and consistent with earlier findings from samples
based on low-resolution spectroscopy \citep[e.g.,][]{Fynbo2009} and from studies
of complete samples of GRB host galaxies \citep{Hjorth2012, Covino2013,
	Perley2016a}. \cite{Kruhler2013} argue, that very rich sightlines like that
probed by the remarkable GRB~080607 \citep{Prochaska2009} are probably
significantly more frequent than in the sightlines sampled by our spectra.
However, with current instrumentation, these sightlines are out of reach except
under very fortunate circumstances as in the case of GRB~080607 when the
afterglow could be observed only a few minutes after the burst with a 10-m class
telescope. Observations of such sightlines with X-shooter-like spectrographs on
the next generation of 20--40-m telescopes is likely to be very rewarding, given
that a suitable GRB detector will be available.

@@

These observations have not been published before. Starting 12.36 hours after
the BAT trigger, 4.8 ks spectroscopic integration yields a very faint trace down
to $\sim$430 nm, only visible after binning heavily. This could indicate a
redshift, $z ~\sim 2.5$, but no emission lines or absorption lines are
immediately visible to support this.

\subsection{GRB~110709B (z=2.109)} \label{110709}

@@

This is a Fermi burst with a LAT detection and thus outside the statistical
sample, but nonetheless followed up due to the extremely high peak energy
\citep{Axelsson2012}. Starting 28.7 hours after the burst trigger, 2.4ks
spectroscopic observation reveals after heavy binning, a wide, faint trace down
to $\sim$ 580 nm, offset by 2.5\arcsec~relative to the centering of the slit.
No good redshift measurement can be inferred from this. We have adopted the
redshift from GCN  12193 \citep{GCN12193}

@@

Starting 6.15 hours after the BAT trigger, spectroscopic integration for 4.8ks
reveals a moderate signal-to-noise GRB afterglow continuum, down to $\sim$ 500 nm. The simultaneous detection of absorption features identified as \lya,
\SIii, \civ, \alii, \cah, \cak, and \mgii, and emission from the \oiii-doublet,
securely sets $z = 3.36$ as the redshift of the GRB. This data forms the basis
of GCN 12284 \citep{GCN12284}, but is not published elsewhere.

@@

This data formed the basis of GCN 12431 \citep{GCN12431} and is additionally
published in \citet{Sparre2014}. Observations of this GRB afterglow was
initiated 7.71 hours after the BAT trigger and had a duration of 8.4ks. A second
observational epoch started 20.1 hours and lasted for 6.6ks. The GRB afterglow
continuum is well detected down to $\sim$760 nm, with several strong
absorption features imprinted, all at a common $z = 4.990$. \lya~is clearly
detected and we additionally detect lines identified as \SIii, \feii, \civ,
\mgii, \SIii*, \sii*, \oi*. An intervening DLA systems is additionally detected
at  $z = 4.61$ as seen from \lya~and \mgii~absorption.

@@

The data presented here has been examined by \citet{Japelj2015} and
Zafar et al. (submitted), who both find a significant amount of extinction, $A_V\approx
1$ mag. Three epochs of observations have been obtained, the first two
immediately after the burst, and the last one long after the afterglow had
faded. Starting 1.4 hours after the \textit{Swift} trigger, the first epoch
contains bright afterglow continuum. Rich in absorption features belonging to a
multitude of ions at $z =    1.728$ is estimated for the host with intervening
systems at $z = 1.476$, $z = 1.214$, $z = 0.662$ and $z = 0.632$. The second
epoch, obtained 4.5 hours after the burst contains the fading afterglow. A third
epoch is obtained $>1$ year after the GRB in which emission lines from \hb~and
\ha~are found at the redshift of the host, confirming the association of the
absorption line system and the host. We also detect C\,\textsc{i} in absorption
which indicates the presence of cold gas.

@@

\subsection{GRB~120224A (z = 1.10)} \label{120224}

@@

\subsection{GRB~120311 (z = 0.350)} \label{120311}

@@

\subsection{GRB~120805A (z = 3.9)} \label{120805}

@@

This spectrum has not previously been published. Starting only 45 minutes after
the BAT trigger, $2\times600$ s spectroscopic integration carried on into the
beginning twilight. Observed at very high airmass with a quickly varying
background, a faint afterglow trace is visible across all arms of X-shooter,
down to ~ 480 nm, which if interpreted as the break due to \lya suggests $z =
2.9$. This supports the approximate redshift suggested in GCN 14898
\citep{GCN14898}.

@@

This data formed the basis for 15330 \citep{GCN15330}, but is not published
elsewhere. Starting $\sim$1.5 days after the Fermi-GBM trigger, 4.5 ks
spectroscopic integration captures a modest signal-to-noise GRB afterglow
continuum all the way down to $\sim$320 nm. Imprinted on the continuum are
absorption features, which we identify as due to \lya, \feii,  \mgii, \mgi~at
the same redshift, which we measure to be $z = 1.874$. From the broad absorption
trough due to \lya, we infer \nh~$= 22.0 \pm 0.3$. This spectrum is taken under
ESO programme ID: 092.D-0056(A) (PI: Rau).

\subsection{GRB~131030A (z = 1.296)}	
\label{131030}

@@

\subsection{GRB~140311A (z = 4.954)}	
\label{140311}
